# Reproducible Research: Peer Assessment 2

Health and Economic Impact of Weather Events in the US
======================================================

Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in
fatalities, injuries, and property damage, and preventing such outcomes to the extent
possible is a key concern.

Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health.


Data Processing
===============

* The data is from a compression types of comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
* The data is downloaded , extracted to the folder and read via CSV reader.
* Some of the data also converted into lower case , converted into dataframe via ddplay , combine via plyr and Preview
* New basic function introduced to convert into exponential transformation


# Basic command line using KnitR ( manual )
library(markdown)
library(knitr)
setwd("D:/Google Drive/Coursera/Assignment 5.1/R/coursera-repdata/project2")
knitr::opts_chunk$set(echo=FALSE, fig.path='D:/Google Drive/Coursera/Assignment 5.1/R/coursera-repdata/project2/figure/', cache=TRUE)
knitr::knit2html("storm.analysis.Rmd", options=c("use_xhtml","smartypants","mathjax","highlight_code", "base64_images"))


* Load all the required library

```{r}
options( warn = -1 )
is.installed <- function(mypkg) is.element(mypkg, installed.packages()[,1]) 
if (is.installed('dplyr') == 'FALSE') {install.packages("dplyr");library(dplyr)} else{suppressMessages(library(dplyr))}
if (is.installed('ggthemes') == 'FALSE') {install.packages("ggthemes");library(ggthemes)} else{suppressMessages(library(ggthemes))}
if (is.installed('scales') == 'FALSE') {install.packages("scales");library(scales)} else{suppressMessages(library(scales))}
if (is.installed('RColorBrewer') == 'FALSE') {install.packages("RColorBrewer");suppressMessages(library(RColorBrewer))} else{library(RColorBrewer)}
if (is.installed('lubridate') == 'FALSE') {install.packages("lubridate");library(lubridate)} else{suppressMessages(library(lubridate))}
if (is.installed('ggplot2') == 'FALSE') {install.packages("ggplot2");library(ggplot2)} else{suppressMessages(library(ggplot2))}
if (is.installed('plyr') == 'FALSE') {install.packages("plyr");library(plyr)} else{suppressMessages(library(plyr))}
if (is.installed('knitr') == 'FALSE') {install.packages("knitr");library(knitr)} else{suppressMessages(library(knitr))}
if (is.installed('lattice') == 'FALSE') {install.packages("lattice");library(lattice)} else{suppressMessages(library(lattice))}
if (is.installed('RCurl') == 'FALSE') {install.packages("RCurl");library(RCurl)} else{suppressMessages(library(RCurl))}
if (is.installed('reshape') == 'FALSE') {install.packages("reshape");library(reshape)} else{suppressMessages(library(reshape))}
if (is.installed('car') == 'FALSE') {install.packages("car");library(car)} else{suppressMessages(library(car))}
if (is.installed('gridExtra') == 'FALSE') {install.packages("gridExtra");library(gridExtra)} else{suppressMessages(library(gridExtra))}
if (is.installed('grid') == 'FALSE') {install.packages("grid");library(grid)} else{suppressMessages(library(grid))}
if (is.installed('xtable') == 'FALSE') {install.packages("xtable");library(xtable)} else{suppressMessages(library(xtable))}
```


* The first step is to read the data into a data frame.

```{r}
options( warn = -1 )
curdir <-getwd()
file.url<-'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(file.url,destfile=paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep=""))
storm <- read.csv(bzfile(paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep="")))
#storm <- read.csv(bzfile("c://repdata%2Fdata%2FStormData.csv.bz2"))
length(unique(storm$EVTYPE))
```

* Get the no. of event types

```{r}
length(unique(storm$EVTYPE))
```

* Converting letters to lower casing

```{r}
event_types <- tolower(storm$EVTYPE)
```

* Replace punctation characters with a space

```{r}
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
```

* Get the unique of event types

```{r}
length(unique(event_types))
```

* Get the casualities

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```

* Find events which causing most injury and death

```{r}
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
fatal_events[, c("EVTYPE", "fatalities")]
injury_events[, c("EVTYPE", "injuries")]
```


* Define function for exponent transformation and apply the transformation

```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}

prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)

```

* Calculating loss by event type

```{r}
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

```

* Removing events with no loss

```{r}
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

Results
===============

* Set the levels in order  and produce 2 types of graph and combined as Top deadly weather

```{r}
library(ggplot2)
library(gridExtra)

p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Top deadly weather events in the US (1950-2011)")	
```

* Set the levels in order  and produce 2 types of graph and combined as Weather costs

```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")
	
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Weather costs to the US economy (1950-2011)")
```
