# Reproducible Research: Peer Assessment 2

Health and Economic Impact of Weather Events in the US
======================================================


Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in
fatalities, injuries, and property damage, and preventing such outcomes to the extent
possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks characteristics of major
storms and weather events in the United States, including when and where they occur, as
well as estimates of any fatalities, injuries, and property damage.

Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health. The second most dangerous
event type is the excessive heat. The economic impact of weather events was
also analyzed. Flash floods and thunderstorm winds caused billions of dollars
in property damages between 1950 and 2011. The largest crop damage caused by
drought, followed by flood and hails.


Data Processing
===============

The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

* Load all the required library

```{r}
options( warn = -1 )
is.installed <- function(mypkg) is.element(mypkg, installed.packages()[,1]) 
if (is.installed('dplyr') == 'FALSE') {install.packages("dplyr");library(dplyr)} else{library(dplyr)}
if (is.installed('ggthemes') == 'FALSE') {install.packages("ggthemes");library(ggthemes)} else{library(ggthemes)}
if (is.installed('scales') == 'FALSE') {install.packages("scales");library(scales)} else{library(scales)}
if (is.installed('RColorBrewer') == 'FALSE') {install.packages("RColorBrewer");library(RColorBrewer)} else{library(RColorBrewer)}
if (is.installed('lubridate') == 'FALSE') {install.packages("lubridate");library(lubridate)} else{library(lubridate)}
if (is.installed('ggplot2') == 'FALSE') {install.packages("ggplot2");library(ggplot2)} else{library(ggplot2)}
if (is.installed('plyr') == 'FALSE') {install.packages("plyr");library(plyr)} else{library(plyr)}
if (is.installed('knitr') == 'FALSE') {install.packages("knitr");library(knitr)} else{library(knitr)}
if (is.installed('lattice') == 'FALSE') {install.packages("lattice");library(lattice)} else{library(lattice)}
if (is.installed('RCurl') == 'FALSE') {install.packages("RCurl");library(RCurl)} else{library(RCurl)}
if (is.installed('reshape') == 'FALSE') {install.packages("reshape");library(reshape)} else{library(reshape)}
if (is.installed('car') == 'FALSE') {install.packages("car");library(car)} else{library(car)}
if (is.installed('gridExtra') == 'FALSE') {install.packages("gridExtra");library(gridExtra)} else{library(gridExtra)}
if (is.installed('grid') == 'FALSE') {install.packages("grid");library(grid)} else{library(grid)}
if (is.installed('xtable') == 'FALSE') {install.packages("xtable");library(xtable)} else{library(xtable)}
```

```r
options( warn = -1 )
is.installed <- function(mypkg) is.element(mypkg, installed.packages()[,1]) 
if (is.installed('dplyr') == 'FALSE') {install.packages("dplyr");library(dplyr)} else{library(dplyr)}
if (is.installed('ggthemes') == 'FALSE') {install.packages("ggthemes");library(ggthemes)} else{library(ggthemes)}
if (is.installed('scales') == 'FALSE') {install.packages("scales");library(scales)} else{library(scales)}
if (is.installed('RColorBrewer') == 'FALSE') {install.packages("RColorBrewer");library(RColorBrewer)} else{library(RColorBrewer)}
if (is.installed('lubridate') == 'FALSE') {install.packages("lubridate");library(lubridate)} else{library(lubridate)}
if (is.installed('ggplot2') == 'FALSE') {install.packages("ggplot2");library(ggplot2)} else{library(ggplot2)}
if (is.installed('plyr') == 'FALSE') {install.packages("plyr");library(plyr)} else{library(plyr)}
if (is.installed('knitr') == 'FALSE') {install.packages("knitr");library(knitr)} else{library(knitr)}
if (is.installed('lattice') == 'FALSE') {install.packages("lattice");library(lattice)} else{library(lattice)}
if (is.installed('RCurl') == 'FALSE') {install.packages("RCurl");library(RCurl)} else{library(RCurl)}
if (is.installed('reshape') == 'FALSE') {install.packages("reshape");library(reshape)} else{library(reshape)}
if (is.installed('car') == 'FALSE') {install.packages("car");library(car)} else{library(car)}
if (is.installed('gridExtra') == 'FALSE') {install.packages("gridExtra");library(gridExtra)} else{library(gridExtra)}
if (is.installed('grid') == 'FALSE') {install.packages("grid");library(grid)} else{library(grid)}
if (is.installed('xtable') == 'FALSE') {install.packages("xtable");library(xtable)} else{library(xtable)}
```


* The first step is to read the data into a data frame.

```r
options( warn = -1 )
curdir <-getwd()
file.url<-'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(file.url,destfile=paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep=""))
storm <- read.csv(bzfile(paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep="")))
```

```{r}
options( warn = -1 )
curdir <-getwd()
file.url<-'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(file.url,destfile=paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep=""))
storm <- read.csv(bzfile(paste(curdir,'/repdata%2Fdata%2FStormData.csv.bz2',sep="")))
```


* Get the no. of event types

```r
length(unique(storm$EVTYPE))
```

```{r}
length(unique(storm$EVTYPE))
```

* Converting letters to lower casing

```r
event_types <- tolower(storm$EVTYPE)
```

```{r}
event_types <- tolower(storm$EVTYPE)
```

* Replace punctation characters with a space

```r
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
```

```{r}
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
```

* Get the unique of event types

```r
length(unique(event_types))
```

```{r}
length(unique(event_types))
```

* Get the casualities

```r
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```

* Find events which causing most injury and death

```r
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
fatal_events[, c("EVTYPE", "fatalities")]
injury_events[, c("EVTYPE", "injuries")]
```

```{r}
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
fatal_events[, c("EVTYPE", "fatalities")]
injury_events[, c("EVTYPE", "injuries")]
```


* Define function for exponent transformation and apply the transformation

```r
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}

prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)

```

```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}

prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)

```

* Calculating loss by event type

```r
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

```

```{r}
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

```

* Removing events with no loss

```r
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```
```{r}
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

* Set the levels in order  and produce 2 types of graph and combined as Top deadly weather

```r
library(ggplot2)
library(gridExtra)

p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")
	
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Top deadly weather events in the US (1950-2011)")	
```
```{r}
library(ggplot2)
library(gridExtra)

p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Top deadly weather events in the US (1950-2011)")	
```

* Set the levels in order  and produce 2 types of graph and combined as Weather costs

```r
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")
	
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Weather costs to the US economy (1950-2011)")	
```
```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")
	
grid.arrange(p1, p2 , ncol=1, nrow=2, top = "Weather costs to the US economy (1950-2011)")
```

